---
title: "ggplot2 for publication + outreach - part 2"
author: "Ryssa Moffat"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes
      smooth_scroll: yes
    theme: spacelab
fontsize: 16pt
---

# Set up
## Loading packages

```{r}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(here, dplyr, ggplot2, ggpubr, viridis, ggh4x, stringr,
               forcats, tidyr)
```


```{r}
# Ryssa's go-to theme
theme_set(theme_light()+
            theme(
              plot.title = element_text(size=rel(1.2),face="bold"),
              axis.title = element_text(size=rel(1),face="bold"),
              axis.text = element_text(size=rel(1),colour = 'black'),
              strip.text = element_text(size=rel(1),colour = 'black', 
                                        face = "bold"),
              legend.text = element_text(size=rel(1)),
              legend.title = element_text(size=rel(1),face="bold"),
              panel.grid = element_blank()))
```


## Data sets
### Robots

We invited kids (5-12 years) to read to different social robots. Afterward, we asked them to rate the robot on a couple attributes.

[Caruana et al. (2022). Perceptions of intelligence & sentience shape children’s interactions with robot reading companions](https://www.nature.com/articles/s41598-023-32104-7).

* `id` participant identifier
* `chosen`which robot the child chose (NAO, Cozmo, MiRo)
* `Enjoy_R` ratings of enjoyment of reading with chosen robot (1-3)
* `Helpful_R` ratings of how helpful chosen robot was during reading (1-3)
* `GoodListener_R` ratings of how chosen robot listened to child reading (1-3)
* `GoodTeacher_R` ratings of chosen robot's skill as teacher (1-3)
* `Kind_R` ratings of chosen robot's kindness (1-3)

```{r}
robot_ratings <- read.csv(here::here("data/robot_ratings.csv")) %>%
  select(-X)

# _R shows it's a rating
# child-friendly ratings scale included: 1 = 'a little bit', 2 = 'somewhere in the middle', 3 = 'a lot'
head(robot_ratings)
```


# 1. Same units = same plot
Let's plot the mean and SD for each rating for each robot.

## Spatially inefficient
Make separate plots and join them up using `ggarrange`. Starting with mean ratings of enjoyment:

```{r}
# get mean value for top of bar plot and sd for error bars in new df
enjoyment_mean <- robot_ratings %>%
  group_by(chosen) %>%
  summarise(mean = mean(Enjoy_R,na.rm=FALSE),
            std = sd(Enjoy_R, na.rm=FALSE))

head(enjoyment_mean)

enjoyment <- enjoyment_mean %>%
  ggplot(aes(x=chosen, y=mean, fill=chosen))+
  geom_bar(stat = "identity", position="dodge")+
  geom_errorbar(aes(ymin=mean-std, ymax=mean+std, group=chosen),
                colour="black", width = 0.25, position=position_dodge(width = 0.9))+
  labs(title = "Ratings of enjoyment", x="", y="Mean rating (1-3)")+
  scale_fill_manual(values=c("lightcoral","lightgoldenrod","lightblue"))+
  guides(fill=guide_legend(title=""))

enjoyment
```

Now ratings of helpfulness:

```{r}
# get mean value for top of bar plot and sd for error bars in new df
helpful_mean <- robot_ratings %>%
  group_by(chosen) %>%
  summarise(mean = mean(Helpful_R,na.rm=FALSE),
            std = sd(Helpful_R, na.rm=FALSE))

helpful <- helpful_mean %>%
  ggplot(aes(x=chosen, y=mean, fill=chosen))+
  geom_bar(stat = "identity", position="dodge")+
  geom_errorbar(aes(ymin=mean-std, ymax=mean+std, group=chosen),
                colour="black", width = 0.25, position=position_dodge(width = 0.9))+
  labs(title = "Ratings of helpulness", x="", y="Mean rating (1-3)")+
  scale_fill_manual(values=c("lightcoral","lightgoldenrod","lightblue"))+
  guides(fill=guide_legend(title=""))

helpful
```

Now joining them up:

```{r}
# plot both
ggarrange(enjoyment, helpful, common.legend = TRUE)
```

## More efficient
To be able to plot the ratings more efficiently, the ratings values all need to be in one columne (instead of separate columns).

```{r}
# get all ratings in one column
rating_long <- robot_ratings %>%
  pivot_longer(cols = ends_with("_R"),
               names_to = "question",
               values_to = "value")

rating_long <- rating_long %>%
  mutate(questionName = str_sub(question,,-3)) #remove _R to improve labels

head(rating_long)

# get mean value for top of bar plot and sd for error bars in new df for all columns
rating_mean <- rating_long %>% 
  group_by(chosen, questionName) %>% 
  summarise(mean = mean(value,na.rm=FALSE),
            std = sd(value,na.rm=FALSE)) %>%
    mutate(chosen = factor(chosen, levels=c("NAO","Cozmo","MiRo")))# robot order

rating_mean %>%
  ggplot(aes(x=questionName, y=mean, fill=chosen))+
  geom_bar(stat = "identity", position="dodge")+
  geom_errorbar(aes(ymin=mean-std, ymax=mean+std, group=chosen),
                colour="black", width = 0.25,
                position=position_dodge(width = 0.9))+
  labs(title = "", x="", y="Mean rating (1-3)")+
  scale_fill_manual(values=c("lightcoral","lightgoldenrod","lightblue"))+
  guides(fill=guide_legend(title=""))+
  facet_grid(.~chosen)+
  guides(color = "none")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  # + coord_flip() # try switching the x and y axis

```

## Spatially efficient

```{r}
rating_mean %>%
  ggplot(aes(x=questionName, y=mean, fill=chosen))+
  geom_bar(stat = "identity", position="dodge")+
  geom_errorbar(aes(ymin=mean-std, ymax=mean+std, group=chosen),
                colour="black", width = 0.25, position=position_dodge(width = 0.9))+
  labs(title = "", x="", y="Mean rating (1-3)")+
  scale_fill_manual(values=c("lightcoral","lightgoldenrod","lightblue"))+
  guides(fill=guide_legend(title=""))
```

## Efficient + fancy

```{r}
rating_mean <- rating_mean %>%
  mutate(questionName =case_when(questionName == "GoodTeacher" ~ "Good\nTeacher",
                                  questionName == "GoodListener" ~ "Good\nListener",
                                 TRUE ~ questionName))
head(rating_mean)

radar_robots <- ggplot(data = rating_mean)+
  geom_hline(aes(yintercept = y), data.frame(y = c(1,2,3)), color = "lightgrey")+
  geom_bar(aes(x=questionName, y=mean, fill=chosen),
           alpha = 0.8, stat = "identity", position="dodge")+
  geom_errorbar(aes(x=questionName,ymin=mean-std, ymax=mean+std, group=chosen),
                colour="black", width = 0.25, alpha = 0.8,
                position=position_dodge(0.9))+
  scale_fill_manual(values=c("lightcoral","lightgoldenrod","lightblue"))+
  labs(title = "Mean rating per robot", x= "", y = "")+
  coord_polar()+
  theme(panel.border = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())

radar_robots +
    annotate(x = .5, y = 3, label = "3", geom = "text")+
    annotate(x = .5, y = 2, label = "2", geom = "text")+
    annotate(x = .5, y = 1, label = "1", geom = "text")
  

```

## All together
Heat maps are surprisingly efficient. Whenever there seem to be too many variables to plot at once, but they have are in the same units, try a heat map (or two).

```{r}
# heatmaps
rating_mean %>%
  ggplot(aes(x=chosen, y=questionName, fill = mean))+
  geom_tile(alpha = 0.9)+
  geom_text(aes(label = round(mean, 1))) +
  scale_fill_viridis() +
  labs(title = "Mean ratings per robot", x="Chosen Robot", y="",
       fill = "Mean rating")

```
