---
title: "ggplot2 for publication + outreach - part 1"
author: "Ryssa Moffat"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes
      smooth_scroll: yes
    theme: spacelab
fontsize: 16pt
---

# Set up
## Loading packages

I love the `pacman` package! It loads and installs the packages as needed, so you never have to use the `library()` and in `install()` functions as often.

The `introdataviz` package can help with raincloud plots. Uncomment it to install it once, then re-comment.

```{r}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(here, # helps keep paths short by starting them from the project
               dplyr, # awesome package for data wrangling
               ggplot2, # the reason we are all here today 
               ggpubr, # helps join up figures in clumps
               ggh4x, # has special function for facetting
               RColorBrewer, # color palette
               viridis, # another color palette
               stringr, # finding/manipulating strings
               tidyr) # great for wrangling

#devtools::install_github("psyteachr/introdataviz") # if you want rainclouds
```


A theme is the canvas behind the data, in a figure. ggplot2 has many basic themes to choose from: [basic themes](https://ggplot2.tidyverse.org/reference/ggtheme.html). They all easily customisable and customisation is your friend! Let's set one here, and come back and play with custom themes once we have made some basic figures. 

```{r}
# Ryssa's go-to theme
theme_set(theme_light()+
            theme(
              plot.title = element_text(size=rel(1.2),face="bold"),
              axis.title = element_text(size=rel(1),face="bold"),
              axis.text = element_text(size=rel(1),colour = 'black'),
              strip.text = element_text(size=rel(1),colour = 'black', 
                                        face = "bold"),
              legend.text = element_text(size=rel(1)),
              legend.title = element_text(size=rel(1),face="bold"),
              panel.grid = element_blank()))
```

## Data sets
### Personality traits

We had 322 people fill in questionnaires about themselves then try to estimate levels of movement synchrony in videos.

[Moffat et al. (2024). Evaluations of dyadic synchrony: observers’ traits influence estimation and enjoyment of synchrony in mirror-game movements](https://www.nature.com/articles/s41598-024-53191-0).

This data frame include the following columns:

* `ID` partcipant identifiers
* `age` participant age
* `extraversion_z` z-scored extraversion scores
* `selfEsteem_z` z-scored self-esteem scores
* `bpq_z` z-scored body perception scores (attention paid to internal bodily functions)
* `bcq_z` z-scored body competence scores (belief in body's ability to complete physical tasks)
* `iri_z` z-scored empathy scores
* `cati_z` z-scored autism trait scores

```{r}
personality_traits <- read.csv(here::here("data/personality_traits.csv")) %>%
  select(-X) # remove column X

#_z after each personality measure shows that the values are z-scored
head(personality_traits)
```


# 1. Aesthetics + colors

Every figure start with the axis and some grouping information. We put these into the `aes()` function.

```{r}
# two different ways to start the same plot
# the dataframe goes right after ggplot()
ggplot(data = personality_traits, aes(x = extraversion_z, y = age))

ggplot(data = personality_traits, aes(x = extraversion_z, y = age))+
  geom_point()


# or you can start with the dataframe and pipe (%>%) into a plot
personality_traits %>%
  ggplot(aes(x = extraversion_z, y = age))+
  geom_point(color = "blue") # try "red", "cornflowerblue", "maroon"

```

## Colors
Many colors can be called by name: [627 ggplot color names](https://www.datanovia.com/en/blog/awesome-list-of-657-r-color-names/). 

They can also be called using hexcodes: [html color picker](https://htmlcolorcodes.com/).

Super useful for matching institutional colors/poster background is that you can match colors from images/screenshots: [html color matcher](https://imagecolorpicker.com/). 

For discrete data (examples below), pre-made color palettes save a lot of time: [Color palettes](https://www.datanovia.com/en/blog/top-r-color-palettes-to-know-for-great-data-visualization/).

```{r}
personality_traits %>%
  ggplot(aes(x = extraversion_z, y = age))+
  geom_point(color = "#dd4e2c", size = 3) # Rladies purple #89398a

```


# 2. Geoms

Geoms refers to the type of plot (e.g., scatter plot, boxplot, barplot, heatmap, etc.). Each field seems to have it's favourites, so here are the one's I use most. When I need inspiration, I peruse [50 best plots with code](https://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html) and look for a geom that's suitable for my data.

In the plots below, try playing with the parameters inside the different `geom_x()` functions.

## Continuous data
```{r}
# distributions (only need x OR y)
# try changing fill, color, linewidth, alpha (0-1)
personality_traits %>%
  ggplot(aes(x = extraversion_z)) +
  geom_density(fill = "lightblue", color = "orange", linewidth = 4, alpha = 1) 


# histograms (only need x OR y)
# try changing fill, color, binwith (in same units as data)
personality_traits %>%
  ggplot(aes(x = extraversion_z)) +
  geom_histogram(fill = "lightblue", color = "purple", binwidth = .2)


# scatter plot
# try changing color, shape, and size
personality_traits %>%
  ggplot(aes(x = extraversion_z, y = age)) +
  geom_jitter(color = "purple", fill = "yellow", shape = 21, size = 3)


# scatter plot with trend line
# try changing color, fill, method (loess, lm)
personality_traits %>%
  ggplot(aes(x = extraversion_z, y = age)) +
  geom_jitter(color = "purple", fill = "salmon", shape = 25, size = 4)+
  geom_smooth(method = "loess", color = "maroon", fill = "darkgreen")


# same plot with vertical and horizontal lines
# try changing linetype (dashed, dotted, solid), and moving the line geoms above the jitter geom
personality_traits %>%
  ggplot(aes(x = extraversion_z, y = age)) +
  geom_jitter(color = "purple", fill = "salmon", shape = 25, size = 4)+
  geom_vline(xintercept = 1, color = "maroon", linetype = "dashed",
             linewidth =2)+
  geom_hline(yintercept = 30, color = "orange", linetype = "solid",
             linewidth =3)

```

## Discrete

First, let's make age into a categorical variable.

```{r}
# make a column for a categorical age variable
personality_traits1 <- personality_traits %>%
  mutate(agegroup = as.factor(case_when(age < 26 ~ "<25",
                                       age > 24 & age < 40 ~ "25-39", 
                                       age > 39 ~ "40+")))

levels(personality_traits1$agegroup)

# get n per group
traits_age_n <- personality_traits1 %>%
  group_by(agegroup) %>%
  tally()
``` 

With discrete variables, we can use "grouping aesthetics".

```{r}
# barplot (for count data)
traits_age_n %>%
  ggplot(aes(x = agegroup, y = n))+
  geom_bar(stat = "identity", fill = "#89398a")
  
# boxplots (group with continuous, base ggplot colors)
personality_traits1 %>%
  ggplot(aes(x = agegroup, y = extraversion_z, fill = agegroup))+
  geom_boxplot()

# violin plot
personality_traits1 %>%
  ggplot(aes(x = agegroup, y = extraversion_z, fill = agegroup))+
  geom_violin() +
  scale_fill_brewer(palette = "Dark2")

# mutliple overlayed geoms
personality_traits1 %>%
  ggplot(aes(x = agegroup, y = extraversion_z, fill = agegroup))+
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey")+
  geom_violin(alpha = 0.7)+
  geom_jitter(width = 0.1, size = 1)+
  scale_fill_brewer(palette = "Dark2")

# mutliple side by side
personality_traits1 %>%
  ggplot(aes(x = agegroup, y = extraversion_z, fill = agegroup))+
  geom_boxplot(alpha = 0.7, width = 0.3, position= position_nudge(x=-.3))+
  geom_jitter(width = 0.1, size = 1)+
  scale_fill_viridis(discrete = TRUE) # add option="magma"

personality_traits1 %>%
  ggplot(aes(x = agegroup, y = extraversion_z, fill = agegroup))+
  introdataviz::geom_flat_violin(position=position_nudge(x=.4),
                                 adjust = 0.5, alpha = 0.8)+
  geom_boxplot(width = 0.1, position=position_nudge(x=.25), alpha = 0.8)+
  geom_point(shape = 21, position = position_jitter(width = .15), alpha = 0.8)+
  scale_fill_viridis(discrete = TRUE) # add option="magma"


```

## 2+ variables

Let's take two of our personality traits. `cati_z` is a measure of autism traits and `iri_z` is a measure of empathy.

```{r}
# make a column for a categorical age variable
personality_traits2 <- personality_traits1 %>%
  select(ID, age, iri_z, cati_z, agegroup) %>%
  pivot_longer(cols = c(iri_z, cati_z),
               names_to = "measure", values_to = "value")

```

```{r}
# dumbel plots / join-up dots
personality_traits2 %>%
  ggplot(aes(x = measure, y = value, group = ID))+
  geom_point()+
  geom_path()+
  facet_grid(.~agegroup)

# let's add a column with whether the difference is positive or negative
personality_traits3 <- personality_traits1 %>%
  mutate(direction = iri_z - cati_z) %>%
  mutate(direction = as.factor(
    case_when(direction > 0 ~ "positive",TRUE ~ "negative"))) %>%
  select(ID, age, cati_z, iri_z, agegroup, direction) %>%
  pivot_longer(cols = c(iri_z, cati_z),
               names_to = "measure", values_to = "value")

# dumbels with colour
personality_traits3 %>%
  ggplot(aes(x = measure, y = value, group = ID, color = direction))+
  geom_point()+
  geom_path()+
  # set colours manually
  scale_color_manual(values = c("blue", "magenta"))+
  # make pannels for each category
  facet_grid(.~agegroup)


```

# 3. Themes

Back to basic `ggplot2` themes and customisation! Play around and change parameters.

## ggplot themes
```{r}
# try changing theme
personality_traits3 %>%
  ggplot(aes(x = measure, y = value, group = ID, color = direction))+
  geom_point()+
  geom_path()+
  # set colours manually
  scale_color_manual(values = c("blue", "magenta"))+
  # make pannels for each category
  facet_grid(.~agegroup)+
  # theme_grey()+
  # theme_bw()+
  # theme_light()+
  # theme_dark()+
  theme_classic()

```

## Custom themes

I've made these quite extreme so you can get a feeling for the different types of parameters that can be customised. A VERY long list of all the individual theme parts you can change is available: [theme components](https://ggplot2.tidyverse.org/reference/theme.html
).

### Example: Too Much

```{r}
theme_set(theme_light()+
            theme(
              plot.background = element_rect(fill = "black"),
              plot.title = element_text(size=rel(2), face="bold", color="lightblue"),
              axis.title = element_text(size=rel(1), face="bold", color="lightblue"),
              axis.text = element_text(size=rel(1), colour='magenta'),
              strip.background = element_rect(fill = "magenta"),
              strip.text = element_text(size=rel(1), colour = 'white', face = "bold"),
              legend.text = element_text(size=rel(1)),
              legend.title = element_text(size=rel(1),face="bold"),
              legend.background = element_rect(fill = "lightblue"),
              panel.background = element_rect(fill = "lightblue"),
              panel.grid.major = element_line(color = "white"),
              panel.grid.minor = element_line(color = "white")))  


personality_traits3 %>%
  ggplot(aes(x = measure, y = value, group = ID))+
  geom_point()+
  geom_path()+
  labs(title = "Palette tester plot", x = "Info", y = "Value")+
  facet_grid(.~agegroup)
```

### Example: Spooky

```{r}
theme_set(theme_light()+
            theme(
              text=element_text(family="mono"),
              plot.background = element_rect(fill = "white"),
              plot.title = element_text(size=rel(2), face="italic"),
              axis.title = element_text(size=rel(1), face="bold", color="black"),
              axis.text = element_text(size=rel(1), colour='black', angle = 45, hjust = 1),
              strip.background = element_rect(fill = "white"),
              strip.text = element_text(size=rel(1), colour = 'orange', face = "bold"),
              legend.text = element_text(size=rel(1)),
              legend.title = element_text(size=rel(1),face="bold"),
              panel.background = element_rect(fill = "white"),
              panel.grid.major.x = element_blank(),
              panel.grid.major.y = element_line(color = "red", linewidth = 3),
              panel.grid.minor = element_line(color = "orange", linetype = "dashed", linewidth = 2))) 

personality_traits3 %>%
  ggplot(aes(x = measure, y = value, group = ID))+
  geom_point()+
  geom_path()+
  labs(title = "Palette tester plot", x = "Info", y = "Value")+
  facet_grid(.~agegroup)
```

### Example: R Ladies
After this plot, practice saving the figure.
```{r}
theme_set(theme_light()+
            theme(
              plot.title = element_text(size=rel(1.2),face="bold"),
              axis.title = element_text(size=rel(1),face="bold"),
              axis.text = element_text(size=rel(1),colour = 'black'),
              strip.text = element_text(size=rel(1),colour = 'white', 
                                        face = "bold"),
              strip.background =element_rect(fill="#89398a"),
              legend.text = element_text(size=rel(1)),
              legend.title = element_text(size=rel(1),face="bold"),
              panel.border = element_rect(color = "#89398a"),
              panel.grid = element_blank(),
              panel.grid.major = element_line(color = "#89398a",
                                              linetype = "dotted", linewidth = 0.5)))


personality_traits3 %>%
  ggplot(aes(x = measure, y = value, group = ID, color = direction))+
  geom_point()+ # try out transparency levels
  geom_path()+
  scale_color_manual(values = c("#38A3A5", "#80ED99"))+
  labs(title = "Palette tester plot", x = "Info", y = "Value")+
  facet_grid(.~agegroup)


# Don't forget to save!
ggsave("figures/RLadies_traits.jpg", height = 3, width = 6, dpi = 400)
```
# 4. Combining plots

```{r}
differences <- personality_traits3 %>%
  ggplot(aes(x = measure, y = value, group = ID, color = direction))+
  geom_point()+ # try out transparency levels
  geom_path()+
  scale_color_manual(values = c("#38A3A5", "#80ED99"))+
  labs(x = "Info", y = "Value")+
  facet_grid(.~agegroup)

distributions <- personality_traits3 %>%
  ggplot(aes(x = value, fill = measure))+
  geom_density(alpha = 0.6)+
  scale_fill_manual(values = c("#361C35", "#89398a"))+
  labs(x = "Value", y = "Density")
```

## Arrangement

```{r}
# top/bottom
ggarrange(distributions, differences, nrow=2)
ggsave("figures/RLadies_combo1.jpg", height = 6, width = 6, dpi = 400)

# side by side
ggarrange(distributions, differences, widths = c(0.6, 1))
ggsave("figures/RLadies_combo2.jpg", height = 3, width = 10, dpi = 400)

```

## Common legends

```{r}
differences <- personality_traits3 %>%
  ggplot(aes(x = measure, y = value,color = measure))+
  geom_jitter(width = 0.2)+ # try out transparency levels
  scale_color_manual(values = c("#38A3A5", "#80ED99"))+
  labs(x = "Info", y = "Value")+
  facet_grid(.~agegroup)

distributions <- personality_traits3 %>%
  ggplot(aes(x = value, fill = measure))+
  geom_density(alpha = 0.6)+
  scale_fill_manual(values = c("#38A3A5", "#80ED99"))+
  labs(x = "Value", y = "Density")

# top/bottom
ggarrange(distributions, differences, nrow=2, common.legend = TRUE)
ggsave("figures/RLadies_combo3_50.jpg", height = 6, width = 6, dpi = 50)
```






