---
title: "ggplot2 for publication + outreach - part 3"
author: "Ryssa Moffat"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes
      smooth_scroll: yes
    theme: spacelab
fontsize: 16pt
---

# Set up
## Loading packages

```{r}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(here, dplyr, ggplot2, ggpubr, viridis, ggh4x, stringr,
               forcats, tidyr, lme4, emmeans, plotly)
```


```{r}
# Ryssa's go-to theme
theme_set(theme_light()+
            theme(
              plot.title = element_text(size=rel(1.2),face="bold"),
              axis.title = element_text(size=rel(1),face="bold"),
              axis.text = element_text(size=rel(1),colour = 'black'),
              strip.text = element_text(size=rel(1),colour = 'black', 
                                        face = "bold"),
              legend.text = element_text(size=rel(1)),
              legend.title = element_text(size=rel(1),face="bold"),
              panel.grid = element_blank()))
```


## Data sets
### Movement synchrony

Pairs of people played the mirror game. We measured how well people were able to mirror each other's spontaneous movements to understand how movement synchrony influences response inhibition and the related patterns of brain activity.

[Moffat et al. (2024). Inhibiting responses under the watch of a recently synchronized peer increases self-monitoring: evidence from functional near-infrared spectroscopy](https://www.nature.com/articles/s41598-023-32104-7).

For this example, we keep the columns:

* `ID` participant identifier
* `leader` whether the participant (P) or helper (C) was leading the game
* `meanSim` the mean level of movement synchrony during the game
* `group` whether the participant belonged to the synchronised or control group

```{r}
poseSimilarity <- read.csv(here::here("data/pose_similarity.csv")) %>%
  filter(ID <500) %>%
  select(-c(cutoff, nFrameskip, nFrames, sdSim)) %>%
  mutate(group = factor(case_when(ID <200 ~ "Synchronised", TRUE ~ "Control")),
         leader = factor(case_when(leader == "C" ~ "Helper",
                                   leader == "P" ~ "Participant")))

head(poseSimilarity)
```

### Synchrony + Traits

While people played the mirror game, we tracked the level of synchrony and the complexity of the mirror game movements. The participants also filled in questionnaires about themselves. We were particularly interested in embodied movement knowledge.One such measure is "body competence", which indexes how strongly people believe that their body can accomplish physical activities.

[Moffat et al. (2024). Dyadic body competence predicts movement synchrony during the mirror game](https://www.frontiersin.org/journals/human-neuroscience/articles/10.3389/fnhum.2024.1401494/full).

For this example, we keep the columns:

* `ID` participant identifier
* `M_similarity` mean movement synchrony
* `M_entropy` mean movement complexity computed using entropy
* `mean_BCQ` mean body competence score per dyad
* `Role` whether the participant was leading the game (leader) or following the helper (following)

We also create a new column:

* `BCQ_half` to categorise dyads as being above or below the median of `mean_BCQ`

## Prepare data

```{r}

poses_traits <- read.csv(here("data/mirrorring_embodiment.csv")) %>%
    select(ID, M_similarity, M_entropy, BCQ, Conf_BCQ, Role, video_code) 

# Calculate mean and z-scores for helper and participant BCQ
poses_traits1 <- poses_traits %>% 
  # Calculate the mean separately for each row (pair)
  rowwise() %>% 
  mutate(mean_BCQ = mean(c(BCQ, Conf_BCQ)), # BCQ = body competence
         BCQ_half = factor(case_when(mean_BCQ > 7.5 ~ "High",
                              TRUE ~ "Low"))) %>%
  select(ID, M_similarity, M_entropy, mean_BCQ, BCQ_half, Role) 

head(poses_traits1)
```


# 1. Model to figure
Let's compare the level of synchrony observed during the mirror game (`group` = Synchronised) vs. the control game (`group` = Control), and find out if it matters who was leading (Leader or Helper in `leader`).

```{r}
head(poseSimilarity)
```


## Model with lme4

```{r}
model <- lmer(meanSim ~ group*leader + (1|ID), data = poseSimilarity)
summary(model)
```


## Estimates with emmeans

```{r}
model %>% emmeans(specs = ~ group + leader)

model_estimates <- model %>% emmeans(specs = ~ group + leader) %>%
  as.data.frame()

head(model_estimates)
```

## Plot estimates
### Geom_point ()

```{r}
model_estimates %>%
  ggplot(aes(x = group, y = emmean, color = leader))+
  geom_point(size = 4, position = position_dodge(0.7))+
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                position = position_dodge(0.7), width = 0.1)+
  scale_color_manual(values = c("#89398a", "#398a89"), name = "Leader")+
  ylim(0,1) +
  labs(x = "Group", y = "Parameter estimate and 95% CI")
```
### Geom_bar+point()

```{r}
estimates <- model_estimates %>%
  ggplot(aes(x = group, y = emmean, fill = leader))+
  geom_bar(stat = "identity", alpha = 0.9, position = position_dodge(.9))+
  geom_point(size = 4, position = position_dodge(.9))+
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                position = position_dodge(.9), width = 0.1)+
  scale_fill_manual(values = c("#89398a", "#398a89"), name = "Leader")+
  ylim(0,1) +
  labs(x = "Group", y = "Parameter estimate and 95% CI")+
  theme(legend.position = c(0.2, 0.85))

estimates
```


## Contrasts with emmeans

```{r}
model %>% emmeans(specs = ~ group) %>% pairs()

group_contrast <- model %>% emmeans(specs = ~ group) %>% 
  pairs() %>%
  as.data.frame()

model %>% emmeans(specs = ~ leader) %>% pairs()

leader_contrast <- model %>% emmeans(specs = ~ leader) %>%
  pairs() %>%
  as.data.frame()

contrast_df <- rbind(group_contrast, leader_contrast)

head(contrast_df)
```
## Plot contrasts

```{r}
contrasts <- contrast_df %>%
  ggplot(aes(x = contrast, y = estimate, fill = contrast))+
  geom_hline(yintercept = 0, color = "grey", linetype = "dashed")+
  geom_errorbar(aes(ymin = estimate-(SE*1.96), #*1.96 for condidence intervals
                    ymax = estimate+(SE*1.96), width = 0.1))+  
  geom_point(size = 4, shape =21)+
  scale_fill_manual(values = c("magenta", "#80ED99"))+
  labs(x = "Contrast estimate", y = "Contrast")+
  theme(legend.position = "none")+
  annotate(x = 1, y = -0.3, label = "***", geom = "text")

contrasts

```

## Combine plots

In the rendered markdown, the labels overlap. This is not the case in the saved image.

```{r}
ggarrange(estimates, contrasts)
ggsave("figures/modelplots.jpg", height = 4, width = 9, dpi = 400)
```



# 2. 3D plots

3D plots can help visualise data with multiple continuous dimensions. The data set with movement synchrony, movement complexity and dyad-level measures of body competence is perfect for this example.

```{r}
head(poses_traits1)
```

## Continuous color scale

The continuous color scale for one axis can help give the plot depth, and thereby help interpretation when rotating it.

```{r}
poses_traits1 %>%
  plot_ly(x = ~M_similarity, y = ~mean_BCQ, z = ~M_entropy,
          marker = list(color = ~M_similarity,
                        colorscale = c('#FFE1A1', '#683531'),
                        showscale = TRUE)) %>%
  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'Synchrony'),
                                   yaxis = list(title = 'Dyadic Body Competence'),
                                   zaxis = list(title = 'Complexity')),
                      annotations = list(
                        x = 1.13,
                        y = 1.05,
                        text = 'Synchrony',
                        xref = 'paper',
                        yref = 'paper',
                        showarrow = FALSE
                        ))
```

## Discrete colors

Alternatively, by creating a new column with which to divide the data into categories, we can help out interpretation too.

```{r}
poses_traits1 %>%
  plot_ly(x = ~M_similarity, y = ~mean_BCQ, z = ~M_entropy,
          color = ~BCQ_half, colors = c("#89398a", "#398a89"), opacity = 0.5) %>%
  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'Synchrony'),
                     yaxis = list(title = 'Dyadic Body Compentence'),
                     zaxis = list(title = 'Complexity')))
       
```



